---
title: OBIS species richness for OSPAR
date: "`r Sys.Date()`"
author: Pieter Provoost
output: (function(...) {
  rmdformats::robobook(toc_depth = 4, pandoc_args = c("+RTS", "-K2000m", "-RTS"), ...) })
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })  
---

## Loading occurrences from parquet and indexing to H3

The biodiversity indicators will be calculated for hexagonal grid cells. Load occurrences from the full parquet export on S3, and index on the H3 hexagonal grids of resolution 3 and 4. 
```{r message=FALSE, warning=FALSE}
library(tibble)
library(arrow)
library(dplyr)
library(sf)
library(glue)
library(h3)

space <- S3FileSystem$create(
  anonymous = TRUE,
  scheme = "https",
  endpoint_override = "ams3.digitaloceanspaces.com"
)

df <- open_dataset(space$path("obis-datasets/exports/obis_20220710.parquet")) %>%
  select(decimalLongitude, decimalLatitude, minimumDepthInMeters, maximumDepthInMeters, species) %>%
  filter(!is.na(species)) %>%
  map_batches(function(batch) {
    batch %>%
      as.data.frame() %>%
      mutate_at(vars(minimumDepthInMeters, maximumDepthInMeters), as.numeric) %>%
      rowwise() %>%
      mutate(depth = mean(na.omit(c(minimumDepthInMeters, maximumDepthInMeters)))) %>%
      select(-minimumDepthInMeters, -maximumDepthInMeters) %>%
      group_by(decimalLongitude, decimalLatitude, depth, species) %>%
      summarize(records = n()) %>%
      ungroup() %>%
      st_as_sf(coords = c("decimalLongitude", "decimalLatitude")) %>%
      mutate(cell3 = geo_to_h3(., res = 3)) %>%
      mutate(cell4 = geo_to_h3(., res = 4)) %>%
      mutate(p3 = substr(cell3, 5, 6)) %>%
      mutate(p4 = substr(cell4, 5, 6)) %>%
      st_drop_geometry()
  })

partitions3 <- unique(df$p3)
partitions4 <- unique(df$p4)
```

### Write to disk

Write the datasets to disk as partitioned parquet files and clean up the data frame:

```{r message=FALSE, warning=FALSE}
temp_dir <- tempdir()
dir3 <- file.path(temp_dir, "cell3")
dir4 <- file.path(temp_dir, "cell4")
unlink(dir3, recursive = TRUE)
unlink(dir4, recursive = TRUE)

df %>% select(-cell4, -p4) %>% write_dataset(path = dir3, partitioning = "p3", format = "parquet")
df %>% select(-cell3, -p3) %>% write_dataset(path = dir4, partitioning = "p4", format = "parquet")

remove(df)
gc()
```

## Calculating indicators

The following method calculates the number of records, species richness, Simpson index, Shannon index, Hurlbert index, and Hill numbers for each cell:

```{r message=FALSE, warning=FALSE}
library(gsl)

calc <- function(df, cell_col = "cell", esn = 50) {
  df <- df %>% rename(cell = !!sym(cell_col))
  t1 <- df %>%
    group_by(cell, species) %>%
    summarize(ni = sum(records))
  t2 <- t1 %>%
    group_by(cell) %>%
    mutate(n = sum(ni))
  t3 <- t2 %>%
    group_by(cell, species) %>%
    mutate(
      hi = -(ni/n*log(ni/n)), si = (ni/n)^2, qi = ni/n,
      esi = case_when(
        n-ni >= esn ~ 1-exp(lngamma(n-ni+1)+lngamma(n-esn+1)-lngamma(n-ni-esn+1)-lngamma(n+1)),
        n >= esn ~ 1
      )
    )
  t4 <- t3 %>%
    group_by(cell) %>%
    summarize(n = sum(ni), sp = n(), shannon = sum(hi), simpson = sum(si), maxp = max(qi), es = sum(esi))
  result <- t4 %>%
    mutate(hill_1 = exp(shannon), hill_2 = 1/simpson, hill_inf = 1/maxp)
  return(result)
}
```

Then read each partition, perform the calculations, and transform to a spatial grid:

```{r message=FALSE, warning=FALSE}
results3 <- purrr::map(partitions3, function(partition_id) {
  ss <- open_dataset(dir3) %>%
    filter(p3 == partition_id) %>%
    collect()
  res <- calc(ss, cell_col = "cell3")
  gc()
  res
}) %>%
  bind_rows() %>%
  mutate(h3_to_geo_boundary_sf(cell)) %>%
  st_as_sf() %>%
  st_wrap_dateline()

results4 <- purrr::map(partitions4, function(partition_id) {
  ss <- open_dataset(dir4) %>%
    filter(p4 == partition_id) %>%
    collect()
  res <- calc(ss, cell_col = "cell4")
  gc()
  res
}) %>%
  bind_rows() %>%
  mutate(h3_to_geo_boundary_sf(cell)) %>%
  st_as_sf() %>%
  st_wrap_dateline()
```

Use the same procedure to calculate results for shallow and deep occurrences only:

```{r message=FALSE, warning=FALSE}
results3_shallow <- purrr::map(partitions3, function(partition_id) {
  ss <- open_dataset(dir3) %>%
    filter(p3 == partition_id & depth < 100) %>%
    collect()
  res <- calc(ss, cell_col = "cell3")
  gc()
  res
}) %>%
  bind_rows() %>%
  mutate(h3_to_geo_boundary_sf(cell)) %>%
  st_as_sf() %>%
  st_wrap_dateline()

results3_deep <- purrr::map(partitions3, function(partition_id) {
  ss <- open_dataset(dir3) %>%
    filter(p3 == partition_id & depth >= 100) %>%
    collect()
  res <- calc(ss, cell_col = "cell3")
  gc()
  res
}) %>%
  bind_rows() %>%
  mutate(h3_to_geo_boundary_sf(cell)) %>%
  st_as_sf() %>%
  st_wrap_dateline()
```

## Maps

```{r message=FALSE, warning=FALSE}
library(rnaturalearth)
library(rnaturalearthdata)
library(viridis)
library(ggplot2)
library(orthoview)
library(scales) 

world <- ne_countries(scale = "medium", returnclass = "sf")
boundary_atlantic <- st_bbox(c(xmin = -45, xmax = 55, ymin = 10, ymax = 90)) %>%
  st_as_sfc() %>%
  st_segmentize(1) %>%
  st_set_crs(4326)
boundary_naces <- read_sf("NACES_MPA_Shp/NACES_MPA.shp")
boundary_ospar <- read_sf("OSPAR_Subregions/OSPAR_subregions_20160418_3857.shp") %>% st_buffer(1000) %>% st_union() %>% st_transform(4326)

lon <- -10
lat <- 40

ortho3 <- results3 %>% orthoview(lon, lat)
ortho4 <- results4 %>% orthoview(lon, lat)
ortho3_shallow <- results3_shallow %>% orthoview(lon, lat)
ortho3_deep <- results3_deep %>% orthoview(lon, lat)
world_ortho <- world %>% orthoview(lon, lat)

make_plot <- function(df, variable, name, boundary, trans = "identity", ...) {
  ggplot() +
    geom_sf(data = df, aes_string(fill = variable), lwd = 0) +
    scale_fill_viridis(option = "inferno", na.value = "white", trans = trans, labels = comma, name = name) +
    geom_sf(data = world_ortho, fill = "#ffffff", color = "#000000", size = 0.1) +
    geom_sf(data = boundary, color = "#ff4000", fill = NA, size = 1.3) +
    theme_minimal() +
    guides(fill = guide_colourbar(barwidth = 10)) +
    theme(panel.grid.major = element_blank(), legend.position = "bottom") +
    coord_sf(crs = create_ortho_string(lon, lat), ...)
}
```

```{r message=FALSE, warning=FALSE}
make_plot(ortho3, "n", "Records", boundary_atlantic, "log10")
```

```{r message=FALSE, warning=FALSE}
make_plot(ortho3, "sp", "Species richness", boundary_atlantic, "log10")
```

```{r message=FALSE, warning=FALSE}
make_plot(ortho3, "n", "Records", boundary_ospar, "log10")
```

```{r message=FALSE, warning=FALSE}
make_plot(ortho3, "sp", "Species richness", boundary_ospar, "log10")
```

```{r message=FALSE, warning=FALSE}
make_plot(ortho4, "n", "Records", boundary_naces, "log10")
```

```{r message=FALSE, warning=FALSE}
make_plot(ortho4, "sp", "Species richness", boundary_naces, "log10")
```

NACES MPA in detail:

```{r message=FALSE, warning=FALSE}
bb <- boundary_naces %>% st_buffer(10) %>% st_transform(create_ortho_string(lon, lat)) %>% st_bbox()
make_plot(ortho3, "sp", "Species richness", boundary_naces, "log10", xlim = c(bb$xmin, bb$xmax), ylim = c(bb$ymin, bb$ymax))
```

```{r message=FALSE, warning=FALSE}
make_plot(ortho4, "sp", "Species richness", boundary_naces, "log10", xlim = c(bb$xmin, bb$xmax), ylim = c(bb$ymin, bb$ymax))
```

### Shallow occurrences

```{r message=FALSE, warning=FALSE}
make_plot(ortho3_shallow, "n", "Records", boundary_ospar, "log10")
```

```{r message=FALSE, warning=FALSE}
make_plot(ortho3_shallow, "sp", "Species richness", boundary_ospar, "log10")
```

### Deep occurrences

```{r message=FALSE, warning=FALSE}
make_plot(ortho3_deep, "n", "Records", boundary_ospar, "log10")
```

```{r message=FALSE, warning=FALSE}
make_plot(ortho3_deep, "sp", "Species richness", boundary_ospar, "log10")
```

## Species lists

```{r message=FALSE, warning=FALSE}
create_species_list <- function(boundary, filename, min_depth = NULL, max_depth = NULL) {
  cells <- polyfill(boundary, res = 4)
  df <- open_dataset(dir4) %>%
    filter(cell4 %in% cells)
  if (!is.null(min_depth)) {
    df <- df %>% filter(depth >= min_depth)
  }
  if (!is.null(max_depth)) {
    df <- df %>% filter(depth < max_depth)
  }
  species <- df %>% distinct(species) %>% collect()
  write.table(species, filename, row.names = FALSE, col.names = FALSE, quote = FALSE)
}
```

All species North Atlantic:

```{r message=FALSE, warning=FALSE}
create_species_list(boundary_atlantic, "output/species_atlantic.txt")
```

All species OSPAR:

```{r message=FALSE, warning=FALSE}
create_species_list(boundary_ospar, "output/species_ospar.txt")
```

All species NACES MPA:

```{r message=FALSE, warning=FALSE}
create_species_list(boundary_naces, "output/species_naces.txt")
```

Shallow species North Atlantic:

```{r message=FALSE, warning=FALSE}
create_species_list(boundary_atlantic, "output/species_atlantic_shallow.txt", max_depth = 100)
```

Shallow species OSPAR:

```{r message=FALSE, warning=FALSE}
create_species_list(boundary_ospar, "output/species_ospar_shallow.txt", max_depth = 100)
```

Shallow species NACES MPA:

```{r message=FALSE, warning=FALSE}
create_species_list(boundary_naces, "output/species_naces_shallow.txt", max_depth = 100)
```

Deep species North Atlantic:

```{r message=FALSE, warning=FALSE}
create_species_list(boundary_atlantic, "output/species_atlantic_deep.txt", min_depth = 100)
```

Deep species OSPAR:

```{r message=FALSE, warning=FALSE}
create_species_list(boundary_ospar, "output/species_ospar_deep.txt", min_depth = 100)
```

Deep species NACES MPA:

```{r message=FALSE, warning=FALSE}
create_species_list(boundary_naces, "output/species_naces_deep.txt", min_depth = 100)
```
